---
title: "RMS Titanic"
author: "Maya Sivanandan"
date: 1-2-2023
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|------------------|-----------------------------|-------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
df_titanic %>% glimpse()
```

**Observations**:

-   Class, Sex, Age, Survived, n (number of)

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   Are there any differences?
    -   Yes, the Wikipedia page references a total of 2,224 people on board (and also 885 crew members and 1,317 passengers which totals to 2,202)
-   If yes, what might account for those differences?
    -   Passengers who bought tickets and didn't actually get on the ship
    -   Passengers travelling under pseudonyms
    -   There might be passengers which we don't know all the information of, so they weren't included in the data set

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>%
  filter(Survived == "Yes") %>% 
  ggplot() +
  geom_col(aes(x = Class, y = n, fill = Sex)) +
  labs(y = "Number of People")
```

**Observations**:

-   The majority of passengers who survived were female, while the majority of surviving crew were male
-   The class with the fewest surviving people was the 2nd Class
-   In the 3rd Class, the ratio of men:women that survived is about 50:50

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>% 
  ggplot() +
  geom_col(aes(x = Class, y = Prop, fill = Sex)) +
  labs(y = "Proportion of People")


df_prop %>%
  filter(Survived == "Yes") %>% 
  ggplot() +
  geom_col(aes(x = Class, y = Prop, fill = Sex), position = "dodge") +
  labs(y = "Proportion of People")
```

**Observations**:

-   Is there anything *fishy* going on in your plot?
    - The first graph is showing that about 300% of the 1st and 2nd class passengers survived, and that over 100% of the 3rd class passengers and crew members survived
    - Even after separating the bars for male and female people, the proportions don't look correct (Not all of the 1st and 2nd class passengers survived)

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(aes(x = Class, y = Prop, fill = Sex), position = "dodge") +
  facet_wrap(~ Age) +
  labs(y = "Proportion of People")

head(df_prop, 50)
```

**Observations**:

-   In terms of adults, twice as many females survived compared to men within the same class. In general, a person's sex was probably the greatest indicator of their chance of survival.

-   There is no proportion for surviving crew children because there were no children working as crew on the Titanic

-   Overall, children were more likely to survive than adults from the same class. This difference is show most prominently when comparing male adults and children. 

-   Children's data show a more expected trend with all of the first and second class children surviving while less than half of the female children and about 25% of the male children in third class survived.

-   Adult women showed a similar trend with the proportion of survivng first and second class women being almost double that of their counterparts in third class.

-   Interestingly, class doesn't necessarily correlate with an adult male passenger's survival rate. While first class passengers were more likely to survive, second class male passengers were the least likely to survive.

-   The proportion of surviving crew member does not seem to be drastically different to that of passengers. The proportion of surviving male crewmates is fairly similar to that of adult male passengers in third class, and the proportion of surviving female crew is essentially the same as adult female 2nd class passengers. However, this is likely because there were significantly fewer female crew than male (23 vs 862)

-   If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
    -   The survivor proportion values are calculated across all 16 'types' of people aboard the Titanic. For example the proportion of surviving male children in 3rd class is 0.729 because it's calculated from a total of how many people who were male AND children AND in 3rd class survived (35 of 48). It does not include those who do not fit all three of those traits (such as adult men in 3rd class). In contrast, the first graphs in question 4 add together the proportions for adults and children of the same class and sex, giving proportions over 1.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
