---
title: "Aluminum Data"
author: "Maya Sivanandan"
date: 2023
output:
  github_document:
    toc: true
prerequisites:
  - e-data03-pivot-basics
---

*Purpose*: When designing structures such as bridges, boats, and planes, the design team needs data about *material properties*. Often when we engineers first learn about material properties through coursework, we talk about abstract ideas and look up values in tables without ever looking at the data that gave rise to published properties. In this challenge you'll study an aluminum alloy dataset: Studying these data will give you a better sense of the challenges underlying published material values.

In this challenge, you will load a real dataset, wrangle it into tidy form, and perform EDA to learn more about the data.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(latex2exp)
```

*Background*: In 1946, scientists at the Bureau of Standards tested a number of Aluminum plates to determine their [elasticity](https://en.wikipedia.org/wiki/Elastic_modulus) and [Poisson's ratio](https://en.wikipedia.org/wiki/Poisson%27s_ratio). These are key quantities used in the design of structural members, such as aircraft skin under [buckling loads](https://en.wikipedia.org/wiki/Buckling). These scientists tested plats of various thicknesses, and at different angles with respect to the [rolling](https://en.wikipedia.org/wiki/Rolling_(metalworking)) direction.

# Loading and Wrangle
<!-- -------------------------------------------------- -->

The `readr` package in the Tidyverse contains functions to load data form many sources. The `read_csv()` function will help us load the data for this challenge.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/stang.csv"

## Load the data
df_stang <- read_csv(filename)
df_stang
```

Note that these data are not tidy! The data in this form are convenient for reporting in a table, but are not ideal for analysis.

### __q1__ Tidy `df_stang` to produce `df_stang_long`. You should have column names `thick, alloy, angle, E, mu`. Make sure the `angle` variable is of correct type. Filter out any invalid values.

*Hint*: You can reshape in one `pivot` using the `".value"` special value for `names_to`.

```{r q1-task}
## TASK: Tidy `df_stang`
df_stang_long <-
  df_stang %>%
    pivot_longer(
      names_to = c(".value", "angle"),
      names_sep = "_",
      values_to = "val",
      cols = c(-thick, -alloy)
  ) %>%
  mutate(angle = as.integer(angle))

df_stang_long <- df_stang_long[df_stang_long$E >= 0, ]


df_stang_long
```

Use the following tests to check your work.

```{r q1-tests}
## NOTE: No need to change this
## Names
assertthat::assert_that(
              setequal(
                df_stang_long %>% names,
                c("thick", "alloy", "angle", "E", "nu")
              )
            )
## Dimensions
assertthat::assert_that(all(dim(df_stang_long) == c(26, 5)))
## Type
assertthat::assert_that(
              (df_stang_long %>% pull(angle) %>% typeof()) == "integer"
            )

print("Very good!")
```

# EDA
<!-- -------------------------------------------------- -->

## Initial checks
<!-- ------------------------- -->

### __q2__ Perform a basic EDA on the aluminum data *without visualization*. Use your analysis to answer the questions under *observations* below. In addition, add your own *specific* question that you'd like to answer about the data---you'll answer it below in q3.

```{r q2-task}
df_stang_long %>% 
  summary()


df_stang_long %>% 
  # group_by(alloy) %>% 
  # group_by(angle) %>% 
  group_by(thick) %>% 
  summarise(count = n())
```

**Observations**:

- Is there "one true value" for the material properties of Aluminum?
  - No, in the data there seems to be some variation in the measurements of `E` and `nu` across the different samples. `E` varies from 10700 to 9900 and `nu` varies from 0.03100 to 0.3310
  - Furthermore, even though elasticity and Poisson's Ratio are considered material properties (constant for a particular alloy), this isn't necessarily true to real life. The 'known' values are more a value that is representative of the dsitribution of possible values (based on the many aspects that can cause variation in these values from how the atoms are arranged to noise in a signal when collecting data digitally).
- How many aluminum alloys are in this dataset? How do you know?
  - One, the only alloy listed is `al_24st`
- What angles were tested?
  - 0, 45, 90 degrees
- What thicknesses were tested?
  - 0.022, 0.032, 0.064, 0.081 inches
- How does Poisson's ratio vary with respect to thickness?
  - It's hard to make any definitive conclusions from the basic EDA.

## Visualize
<!-- ------------------------- -->

### __q3__ Create a visualization to investigate your question from q2 above. Can you find an answer to your question using the dataset? Would you need additional information to answer your question?

```{r q3-task}
# mid<-mean(df_q2$Temp)

df_stang_long %>%
  ggplot() +
    geom_count(mapping = aes(x = thick, y = nu,
                             color = after_stat(n), size = 1)) +
  
    guides(color = 'legend') +
  
    guides(size = FALSE) +
  
    labs(x = "Thickness (in)", y = TeX("Poisson's Ratio"), 
         title = "Poisson's Ratio vs Thickness")
```

**Observations**:

- Poisson's ratio does not seem to vary with thickness. For the three smallest thicknesses, the majority of the measurements are between 0.317 and 0.330

- The plot does show that the measurements of the 0.081in samples showed slightly lower Poisson ratios, so there may be some sort of relationship between Poisson's Ratio and thickness (perhaps something similar to a negative quadratic where the trend seems initially linear before decreasing). However, we only have the 0.081in measurements showing this downward trend, so this plot can not conclusively support the claim that Poisson's Ratio decreases with increased thickness of a material.


### __q4__ Consider the following statement:

"A material's property (or material property) is an intensive property of some material, i.e. a physical property that does not depend on the amount of the material."[2]

Note that the "amount of material" would vary with the thickness of a tested plate. Does the following graph support or contradict the claim that "elasticity `E` is an intensive material property." Why or why not? Is this evidence *conclusive* one way or another? Why or why not?

```{r q4-vis}
## NOTE: No need to change; run this chunk
df_stang_long %>%

  ggplot(aes(nu, E, color = as_factor(thick))) +
  geom_point(size = 3) +
  theme_minimal()
```

**Observations**:

- Does this graph support or contradict the claim above?
  - When first looking at the graph, it does seem to contradict the claim above, since the measurements of the thickest alloy samples (0.081in) are distinctly different from the others, existing in a separate cluster of data points with lower elasticity and Poisson's ratio. However, when reading the paper on how the measurements were collected, we find that the 0.081in samples were measured on a different apparatus than the other, thinner samples. This could be a source of systematic error, but more tests where all measurements are collected with the same piece of equipment (or all samples are measured on both apparatuses and compared) would be needed to determine if this is true. With the current data, this graph is not conclusive evidence disproving the claim.
  


**Graphs for Group Discussion**
```{r presentation-graph-0} 
df_stang_edit <- 
  df_stang_long[df_stang_long$thick < 0.081, ]
  

df_stang_long %>%
  ggplot() +
    geom_count(mapping = aes(x = thick, y = E)) +
  
    geom_smooth(mapping = aes(x = thick, y = E), method = "lm") +
  
    geom_smooth(df_stang_edit, mapping = aes(x = thick, y = E, color = "red"),
                method = "lm") +
  
    guides(color = FALSE) +
  
    labs(x = "Thickness (in)", y = TeX("Young's Modulus (Kips/ in $^ 2$)"), 
         title = "Young's Modulus vs Thickness")
      
```

```{r presentation-graph-1} 
df_stang_long %>%
  ggplot() +
    geom_count(mapping = aes(x = thick, y = E)) +
  
    geom_smooth(mapping = aes(x = thick, y = E)) +
  
    geom_smooth(df_stang_edit, mapping = aes(x = thick, y = E, color = "red")) +
  
    guides(color = FALSE) +
  
    labs(x = "Thickness (in)", y = TeX("Young's Modulus (Kips/ in $^ 2$)"), 
         title = "Young's Modulus vs Thickness")
```


# References
<!-- -------------------------------------------------- -->

[1] Stang, Greenspan, and Newman, "Poisson's ratio of some structural alloys for large strains" (1946) Journal of Research of the National Bureau of Standards, (pdf link)[https://nvlpubs.nist.gov/nistpubs/jres/37/jresv37n4p211_A1b.pdf]

[2] Wikipedia, *List of material properties*, accessed 2020-06-26, (link)[https://en.wikipedia.org/wiki/List_of_materials_properties]
