---
title: "Gapminder"
author: "Maya Sivanandan"
date: 2023
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(gapminder)


```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an
independent organization that seeks to educate people about the state of the
world. They seek to counteract the worldview constructed by a hype-driven media
cycle, and promote a "fact-based worldview" by focusing on data. The dataset
we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### __q0__ Perform your "first checks" on the dataset. What variables are in this dataset?

```{r q0-task}
gapminder %>% head()

gapminder %>% glimpse()

gapminder %>% summary()

```

**Observations**:

- country, continent, year, lifeExp, pop, gdpPercap

### __q1__ Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- 
  gapminder %>% 
    pull(year) %>% 
      max()

year_min <-
  gapminder %>% 
    pull(year) %>% 
      min()
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### __q2__ Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
df_q2 <- 
  gapminder[gapminder$year == year_min, ]


# df_q2 %>% 
#   ggplot() +
#   aes(continent, gdpPercap)+
#   geom_boxplot()
# 
# 
# df_q2_mod <- df_q2[df_q2$gdpPercap < 90000, ]
# 
# df_q2_mod %>%
#   ggplot() +
#   aes(continent, gdpPercap)+
#   geom_boxplot()


# df_q2 %>% 
#   ggplot() +
#   aes(continent, gdpPercap)+
#   geom_boxplot() +
#   ylim(0, 15000)


df_q2 %>%
  ggplot() +
  aes(continent, gdpPercap)+
  geom_boxplot() +
  scale_y_continuous(trans='log10')


## Check the mean gdpPerCap of Asia with and without Kuwait
# df_q2[df_q2$continent == "Asia", ] %>% 
# summarize(mean(gdpPercap))
# 
# df_q2_mod[df_q2_mod$continent == "Asia", ] %>% 
# summarize(mean(gdpPercap))
```

**Observations**:

- Oceania seems to have the highest `gdpPercap` and the least variation
  - This is likely because in this dataset, the continent of Oceania only has two countries, Australia and New Zealand.
- Europe has the next highest `gdpPerCap`, but it also has the most variation (shown by the lower quartile being at around 3000 and the upper quartile at around 7500) despite not having as having fewer countries than both Asia and Africa (30 vs 33 and 52, respectively)
  

**Difficulties & Approaches**:

- In this dataset, many of the values of `gdpPercap` are similar enough to overlap on the graphy, but they are not the exact same values, so `geom_point()` and `geom_count()` do not best represent the data. I instead chose to use a box plot to show the general trends of the data (mean value as well as the interquartile range)
- When looking at only data from 1952, the `gdpPercap` of Kuwait is over 7 times greater than the next highest GDP (about 108,000 compared to 15,000). This causes all but one of the data points to be represented in the lower 1/8th of the graph (vertically), meaning the points overlap visually and it is hard to make observations.
  - To solve this, I considered changing the limit of the y-axis, such that the data point is not shown visually but is still used to calculate the box plot. Sadly, I don't think it worked as it gives the warning *"Warning: [38;5;252mRemoved 1 rows containing non-finite values (`stat_boxplot()`).[39m"* and the plot looks identical to one I generated where I removed the datapoint before plotting. This does drastically change the boxplot for Asia since the mean `gdpPercap` goes from 5195.484 with Kuwait to 1970.894 without. 
  - I then tried transforming the y-axis to a logarithmic scale, which does show all the data points more evenly spread across the y-axis.Howvever, it becomes harder to visually understand the box plots, since each tick on the y-axis of the graph is a drastically different amount. 

### __q3__ You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
df_q2 %>% 
  ggplot() +
  aes(y = gdpPercap)+
  geom_boxplot() +
  ylim(0, 15000)


df_q2_upper = 3913.5 + (IQR(df_q2$gdpPercap)*1.5)
## No lower outliers


df_q3_o1 <- df_q2[df_q2$gdpPercap > df_q2_upper, ]
df_q3_o1

# df_q2_asia <- df_q2[df_q2$continent == "Asia", ]
# df_q2_upper_asia = 3035.3 + (IQR(df_q2_asia$gdpPercap)*1.5)
# 
# df_q2_americas <- df_q2[df_q2$continent == "Americas", ]
# df_q2_upper_americas = 3940 + (IQR(df_q2_asia$gdpPercap)*1.5)
# 
# df_q2_africa <- df_q2[df_q2$continent == "Africa", ]
# df_q2_upper_africa = 1454.9 + (IQR(df_q2_asia$gdpPercap)*1.5)
# 
# df_q2_europe <- df_q2[df_q2$continent == "Europe", ]
# df_q2_upper_africa = 7236.8 + (IQR(df_q2_asia$gdpPercap)*1.5)
# 
# df_q2_africa%>% 
#   summary()
# 
# 
# df_q3_o2 <- full_join(df_q2_asia[df_q2_asia$gdpPercap > df_q2_upper_asia, ], 
#                       df_q2_africa[df_q2_africa$gdpPercap > df_q2_upper_africa, ])

```

**Observations**:

- Identify the outlier countries from q2
  - According to `geom_boxplot()`, outlying values are those which are more than (1.5 * IQR) below the lower quartile or more than (1.5 * IQR) above the upper quartile. This means that when looking at df_q2, the following countries are outliers: Australia, Bahrain, Canada, Denmark, Kuwait, Netherlands, New Zealand, Norway, Sweden, Switzerland, United Kingdom, United States.
  - This differs from looking at outliers with regard to continent. For example, Australia and New Zealand wouldn't be considered outliers for Oceania

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### __q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent

gapminder %>%
  filter(year == max(year) |  year == min(year)) %>%

  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot()+
  facet_wrap(~year) +
  scale_y_continuous(trans='log10') +
  geom_point(
  data = . %>% 
      group_by(year) %>%
      filter(
          gdpPercap < quantile(gdpPercap, 0.25)- IQR(gdpPercap*1.5) | 
          gdpPercap > quantile(gdpPercap, 0.75)+ IQR(gdpPercap*1.5)
      ),
      mapping = aes(color = country),
      size = 2
)

```

**Observations**:

- Overall, `gdpPercap` increased between 1952 and 2007 for all continents. 
- The United States was an outlier in 1952 and 2007
- The upper quartile for Europe and Americas grew significantly between 1952 and 2007
- The range between New Zealand and Australia grew slightly between 1952 and 2007 (the boxplot for Oceania is not just a line)

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### __q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
gapminder %>%
  filter(year == max(year) |  year == min(year)) %>%

  ggplot(aes(lifeExp, pop)) +
  scale_y_continuous(trans='log10') +
  geom_point(mapping = aes(color = continent), size = 2)+
  facet_wrap(~year) +
  facet_grid(vars(continent), vars(year))

```

- The life expectancy for all continents has increased between 1952 and 2007
- The life expectancy for Europe and Americas seems to have become more concentrated between 1952 and 2007, there is less variation between the different countries
- The life expectancy for countries in Africa has increased overall, but there is also more variation between the countries.
- There is one country in Asia from 2007 that seems to have an outlying life expectancy (~44 years)

```{r q5-task2}
gapminder %>%
  filter(year == max(year) |  year == min(year)) %>%

  ggplot(aes(lifeExp, gdpPercap)) +
  scale_y_continuous(trans='log10') +
  geom_point(mapping = aes(color = continent), size = 2)+
  facet_wrap(~year) +
  facet_grid(vars(continent), vars(year))
```

- Overall, there seems to be a general trend where countries with a higher GDP per capita also have a higher life expectancy
- Between 1952 and 2007, the life expectancy and GDP per capita for the countries in Asia, Americas, and Europe all showed a significant increase
- Europe 2007 seems to show the least variation of life expectancy/GDP between countries (the country points are more clumped together)

```{r q5-task3}
gapminder %>%
  filter(year == max(year) |  year == min(year)) %>%
  ggplot(aes(gdpPercap, pop)) +
  scale_y_continuous(trans='log10') +
  scale_x_continuous(trans='log10') +
  geom_point(mapping = aes(color = continent), size = 2)+
  facet_wrap(~year) +
  facet_grid(vars(continent), vars(year))
```

- Population doesn't seem to vary significantly between 1952 and 2007
- There doesn't seem to be a clear relationship between population and GDP per capita
  - Asia in 1952 shows a slight trend where lower populations had higher GDP per capita
