---
title: "Antibiotics"
author: "Maya Sivanandan"
date: 2023
output:
  github_document
---

*Purpose*: Creating effective data visualizations is an *iterative* process; very rarely will the first graph you make be the most effective. The most effective thing you can do to be successful in this iterative process is to *try multiple graphs* of the same data.

Furthermore, judging the effectiveness of a visual is completely dependent on *the question you are trying to answer*. A visual that is totally ineffective for one question may be perfect for answering a different question.

In this challenge, you will practice *iterating* on data visualization, and will anchor the *assessment* of your visuals using two different questions.

*Note*: Please complete your initial visual design **alone**. Work on both of your graphs alone, and save a version to your repo *before* coming together with your team. This way you can all bring a diversity of ideas to the table!

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(ggrepel)
```

*Background*: The data[1] we study in this challenge report the [*minimum inhibitory concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration) (MIC) of three drugs for different bacteria. The smaller the MIC for a given drug and bacteria pair, the more practical the drug is for treating that particular bacteria. An MIC value of *at most* 0.1 is considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin, streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized into a genus based on a number of features, including their resistance to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()

df_antibiotics
```

# Visualization
<!-- -------------------------------------------------- -->

### __q1__ Prototype 5 visuals

To start, construct **5 qualitatively different visualizations of the data** `df_antibiotics`. These **cannot** be simple variations on the same graph; for instance, if two of your visuals could be made identical by calling `coord_flip()`, then these are *not* qualitatively different.

For all five of the visuals, you must show information on *all 16 bacteria*. For the first two visuals, you must *show all variables*.

*Hint 1*: Try working quickly on this part; come up with a bunch of ideas, and don't fixate on any one idea for too long. You will have a chance to refine later in this challenge.

*Hint 2*: The data `df_antibiotics` are in a *wide* format; it may be helpful to `pivot_longer()` the data to make certain visuals easier to construct.

#### Visual 1 (All variables)

In this visual you must show *all three* effectiveness values for *all 16 bacteria*. You must also show whether or not each bacterium is Gram positive or negative.

```{r q1.1}
# WRITE YOUR CODE HERE
# summary(df_antibiotics)
# head(df_antibiotics)


df_antibiotics_long <-
  df_antibiotics %>%
    pivot_longer(
      cols = ends_with("in"),
      names_to = "drug",
      values_to = "MIC")

# df_antibiotics_long

df_antibiotics_long %>%
  ggplot(aes(bacteria, MIC)) +
  scale_y_continuous(trans='log10') +
  geom_point(mapping = aes(color = drug, shape = gram), size = 2) +
  # coord_flip()
  theme(axis.text.x = element_text(angle = 320, hjust = 0))

```

#### Visual 2 (All variables)

In this visual you must show *all three* effectiveness values for *all 16 bacteria*. You must also show whether or not each bacterium is Gram positive or negative.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.2}
df_antibiotics_long %>% 
  ggplot(aes(drug, bacteria)) +
  scale_fill_gradient(trans="log10", low = "purple", high = "green") +
  geom_tile(mapping = aes(fill = MIC)) +
  geom_text(aes(label = gram))
```

#### Visual 3 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.3}
df_antibiotics_long %>%
  # filter(drug == "neomycin") %>% 
  ggplot(aes(bacteria, MIC)) +
  scale_y_continuous(trans='log10') +
  geom_point(mapping = aes(color = gram), size = 2) +
  coord_flip() +
  facet_wrap(vars(drug))
```

#### Visual 4 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.4}
df_antibiotics_long %>%
  ggplot(aes(drug, MIC)) +
  geom_boxplot() +
  geom_point(mapping = aes(color = gram)) +
  scale_y_continuous(trans='log10')
```

#### Visual 5 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.5}
df_antibiotics_long %>%
  ggplot(aes(bacteria, gram)) +
  geom_point(mapping = aes(color = MIC), size = 2) +
  scale_color_gradient(trans="log10", low = "blue", high = "red") +
  coord_flip() +
  facet_wrap(vars(drug))
```

### __q2__ Assess your visuals

There are **two questions** below; use your five visuals to help answer both Guiding Questions. Note that you must also identify which of your five visuals were most helpful in answering the questions.

*Hint 1*: It's possible that *none* of your visuals is effective in answering the questions below. You may need to revise one or more of your visuals to answer the questions below!

*Hint 2*: It's **highly unlikely** that the same visual is the most effective at helping answer both guiding questions. **Use this as an opportunity to think about why this is.**

#### Guiding Question 1

> How do the three antibiotics vary in their effectiveness against bacteria of different genera and Gram stain?

*Observations*
- What is your response to the question above?
  - Pennicilin seems to be most effective against bacteria that present a negative gram strain, and least effective to those with a positive gram stain
  - Neomycin and streptomycin seem to have similar effects, though streptomycin is more effective against staphylooccocus
- Which of your visuals above (1 through 5) is **most effective** at helping to answer this question?
  - I found visual 5 to be the most helpful for this question. 
- Why?
  - Though the color scale isn't great if I want to read specific MIC values, the difference between red and blue is relatively easy for me to detect. Also having the gram stain results physically separated from each other made it easier to notice patterns
  
#### Guiding Question 2

In 1974 *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*, and in 1984 *Streptococcus fecalis* was renamed *Enterococcus fecalis* [2]. 

> Why was *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*?

*Observations*
- What is your response to the question above?
  - The way the bacteria reacted to the 3 mentioned drugs is similar to other strains of Streptococcus 
- Which of your visuals above (1 through 5) is **most effective** at helping to answer this question?
  - Graph 3
- Why?
  - I could see Diplococcus pneumoniae had points that were in line with the other Streptococcus for all three drugs because all the points were on the same axis


# References
<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." *American Scientist* (2009) [link](https://www.americanscientist.org/article/thats-funny)
